Automated Assessment of Residual Plots with Computer Vision Models

### 2025-07-11 appendix.tex
recompile paper + appendix

\author{\name{Weihao Li$^{a, b}$, Dianne Cook$^{a}$, Emi Tanaka$^{a, b,
c}$, Susan VanderPlas$^{d}$, Klaus Ackermann$^{a}$}
\affil{$^{a}$Department of Econometrics and Business Statistics, Monash
University, Clayton, VIC, Australia; $^{b}$Biological Data Science
Institute, Australian National University, Acton, ACT,
Australia; $^{c}$Research School of Finance, Actuarial Studies and
Statistics, Australian National University, Acton, ACT,
Australia; $^{d}$Department of Statistics, University of Nebraska,
Lincoln, Nebraska, USA}
}

>>>>

\author{\name{Weihao Li$^{a, b}$, Dianne Cook$^{a}$, Emi
Tanaka$^{b}$, Susan VanderPlas$^{c}$, Klaus Ackermann$^{a}$}
\affil{$^{a}$Department of Econometrics and Business Statistics, Monash
University, Clayton, VIC, Australia; $^{b}$Research School of Finance,
Actuarial Studies and Statistics, Australian National University, Acton,
ACT, Australia; $^{c}$Department of Statistics, University of Nebraska,
Lincoln, Nebraska, USA}
}

### 2025-07-11 paper.tex
recompile paper + appendix

\begin{keywords}
statistical graphics; data visualization; visual inference; computer
vision; machine learning; hypothesis testing; reression analysis;
cognitive perception; simulation; practical significance
\end{keywords}

@@

The concept of visual inference, as proposed by
\citet{buja2009statistical}, provides an inferential framework to assess
whether residual plots indeed contain visual patterns inconsistent with
the model assumptions. The fundamental idea involves testing whether the
true residual plot visually differs significantly from null plots, where
null plots are plotted with residuals generated from the residual
rotation distribution \citep{langsrud2005rotation}, which is a
distribution consistent with the null hypothesis \(H_0\) that the linear
regression model is correctly specified. Typically, the visual test is
accomplished through the lineup protocol, where the true residual plot
is embedded within a lineup alongside several null plots. If the true
residual plot can be distinguished from the lineup, it provides evidence
for rejecting \(H_0\).

@@

Deep learning models are sensitive to input design, and several
alternatives are available for encoding residual plots.

A simple approach involves feeding vectors of residuals and fitted
values into the model. While this format contains all relevant
information, the input length varies with sample size, which conflicts
with the fixed-shape requirements of modern vision models such as those
implemented in TensorFlow \citep{abadi2016tensorflow}. Padding can be
used to enforce uniform length but risks truncation if inputs exceed the
fixed size. Alternatively, fixed-size sampling of residual--fitted value
pairs can ensure consistency, though at the cost of some information
loss.

@@

Using multiple residual plots as input, such as pairs, triplets, or full
lineups is also a possible option. For instance, triplet networks
\citep{chopra2005learning} can assess visual similarity if we assume
null plots are more similar to each other than to the true residual
plot, which may help distinguish null plots from true residual plots.
However, these architectures introduce complexity in training due to
weight sharing and specialized loss functions. We experimented with this
setting in a pilot study, and found multiple residual plots led to high
input resolution, high computational cost and suboptimal model
performance.

Balancing accuracy, interpretability, and implementation cost, we adopt
a single residual plot in image format as the model input.

@@

Figure \ref{fig:model-performance} shows a hexagonal heatmap of
\(D - \hat{D}\) versus \(D\). Smoothing curves (fitted by generalized
additive models \citep{hastie2017generalized}) reveal that all models
performed well for \(1.5 < D < 6\), where no structural issues were
observed. Over-prediction occurred when \(D < 1.5\), and
under-prediction primarily when \(\hat{D} > 6\).

@@

In the experiment conducted by \citet{li2024plot}, participants could
select multiple plots per lineup. To reflect detection confidence, a
weighted detection rate was computed for each lineup: selecting zero
plots yielded a weight of 0.05; otherwise, if the true residual plot was
selected, its weight was \(1 / \text{(number of selections)}\).

>>>>

\begin{keywords}
statistical graphics; data visualization; visual inference; computer
vision; machine learning; hypothesis testing; regression analysis;
cognitive perception; simulation; practical significance
\end{keywords}

@@

The concept of visual inference, as proposed by
\citet{buja2009statistical}, provides an inferential framework to assess
whether residual plots indeed contain visual patterns inconsistent with
the model assumptions. The fundamental idea involves testing whether the
true residual plot visually differs significantly from the null plots,
where null plots are plotted with residuals generated from the residual
rotation distribution \citep{langsrud2005rotation}, which is a
distribution consistent with the null hypothesis \(H_0\) that the linear
regression model is correctly specified. Typically, the visual test is
accomplished through the lineup protocol, where the true residual plot
is embedded within a lineup alongside several null plots. If the true
residual plot can be distinguished from the lineup, it provides evidence
for rejecting \(H_0\).

@@

Deep learning models are sensitive to input design, and several
alternatives are available for encoding residual plots. A simple
approach involves feeding vectors of residuals and fitted values into
the model. While this format contains all relevant information, the
input length varies with sample size, which conflicts with the
fixed-shape requirements of modern vision models such as those
implemented in TensorFlow \citep{abadi2016tensorflow}. Padding can be
used to enforce uniform length but risks truncation if inputs exceed the
fixed size. Alternatively, fixed-size sampling of residual--fitted value
pairs can ensure consistency, though at the cost of some information
loss.

@@

Using multiple residual plots as input, such as pairs, triplets, or full
lineups is also a possible option. For instance, triplet networks
\citep{chopra2005learning} can assess visual similarity if we assume
null plots are more similar to each other than to the true residual
plot, which may help distinguish null plots from true residual plots.
However, these architectures introduce complexity in training due to
weight sharing and specialized loss functions. We experimented with this
setting in a pilot study, and found multiple residual plots led to high
input resolution, high computational cost and suboptimal model
performance. Balancing accuracy, interpretability, and implementation
cost, we adopt a single residual plot in image format as the model
input.

@@

Figure \ref{fig:model-performance} shows a hexagonal heatmap of
\(D - \hat{D}\) versus \(D\). Smoothing curves \citep[fitted by
generalized additive models,][]{hastie2017generalized} reveal that all
models performed well for \(1.5 < D < 6\), where no structural issues
were observed. Over-prediction occurred when \(D < 1.5\), and
under-prediction primarily when \(\hat{D} > 6\).

@@

In the experiment conducted by \citet{li2024plot}, participants could
select multiple plots per lineup. To reflect detection confidence, a
weighted detection rate was computed for each lineup: selecting zero
plots yielded a weight of 0.05; otherwise, if the true residual plot was
selected, its weight was the inverse of the number of selections.
